---
title: "Climate Exploration - Ann Arbor, MI"
author: "Kiera Matiska"
date: "2023-04-12"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(lubridate)
library(janitor)
library(Kendall)
library(zoo)
library(kableExtra)
library(scales)
library(dplyr)
```

## Read in Data

```{r}
climate_clean <- read_csv(here("data", "A2_climate_data.csv")) %>% 
  clean_names()

climate <- climate_clean %>% 
  mutate(date = ymd(date)) %>% 
  mutate(year = substr(date, 1, 4)) %>% 
  mutate(month = substr(date, 6, 7)) %>% 
  mutate(day = substr(date, 9, 10)) %>% 
  mutate(season = case_when(
    month == "01" ~ "winter",
    month == "02" ~ "winter",
    month == "03" ~ "winter",
    month == "04" ~ "spring",
    month == "05" ~ "spring",
    month == "06" ~ "spring",
    month == "07" ~ "summer",
    month == "08" ~ "summer",
    month == "09" ~ "summer",
    month == "10" ~ "fall",
    month == "11" ~ "fall",
    month == "12" ~ "fall"
  ))
```

# Exploration Plots {.tabset}

## Precipitation

```{r}
ggplot(climate, 
       aes(x = date, y = prcp)) +
  geom_line()
```

## Snow

```{r}
ggplot(climate,
       aes(x = date, y = snow)) +
  geom_line()
```

## Maximum Temperature

```{r}
ggplot(climate, 
       aes(x = date, y = tmax)) +
  geom_line()
```

## Minimum Temperature

```{r}
ggplot(climate, 
       aes(x = date, y = tmin)) +
  geom_line()
```

```{r, include=FALSE}
# Fill in missing data

# precipitation and snow
climate$prcp[is.na(climate$prcp)] <- 0

climate$snow[is.na(climate$snow)] <- 0

# Maximum and Minimum Temperature
climate$tmax <- na.locf(na.locf(climate$tmax), fromLast = TRUE)

climate$tmin <- na.locf(na.locf(climate$tmin), fromLast = TRUE)

# Add precipitation and snow together for all precipitation
climate <- climate %>% 
  mutate(all_prcp = prcp + snow)
```

# Yearly Trends {.tabset}

## Precipitation (Snow and Rain) {.tabset}

```{r, include=FALSE}
climate_prcp_year_sum <- climate %>% 
  group_by(year) %>% 
  summarize(most_rain = max(prcp),
            least_rain = min(prcp),
            total_rain = sum(prcp),
            most_snow = max(snow),
            least_snow = min(snow),
            total_snow = sum(snow),
            most_all_prcp = max(all_prcp),
            least_all_prcp = min(all_prcp),
            total_all_prcp = total_snow + total_rain) %>% 
  mutate(year = as.numeric(year))
```

### Over Total Time Period {.tabset}

#### Linear Model

```{r}
# ggplot(climate_prcp_year_sum, 
#        aes(x = year, y = total_all_prcp)) +
#   geom_point(color = "blue") +
#   geom_smooth(formula = y ~ x,
#               color = "black",
#               method = "lm",
#               se = FALSE)

# trend line slope
precip_lm <- lm(total_all_prcp ~ year, data = climate_prcp_year_sum)
precip_confint <- confint(precip_lm, "year", level = 0.95)

ggplot(data = climate_prcp_year_sum,
       aes(x = year, y = total_all_prcp)) +
  stat_summary(fun = "mean", 
               geom = "point", 
               color = "blue") +
  geom_smooth(method = "lm",
              color = "black",
              se = FALSE) +
  labs(title = "Total Annual Precipitation by Year from 1891-2023",
       x = "Year",
       y = "Total Annual Precipitation (inches)") +
  theme(plot.title = element_text(hjust = 0.5))
```

The y-intercept value stands at `r round(precip_lm$coefficients[1], 2)` inches. The slope for precipitation by year is `r round(precip_lm$coefficients[2], 3)` and is statistically significant with a p-value of 2.09e-09.

#### Mann-Kendall

```{r}
prcp_mk <- MannKendall(climate_prcp_year_sum$total_all_prcp)
```

The strength of the trend is 0.343 and is statistically significant with a p-value of 2.22e-16.

### Between Time Periods {.tabset}

```{r, include=FALSE}
climate_prcp_91_24 <- climate_prcp_year_sum %>% 
  filter(year %in% 1891:1924) %>% 
  mutate(year = as.numeric(year))

climate_prcp_25_58 <- climate_prcp_year_sum %>% 
  filter(year %in% 1925:1958)

climate_prcp_59_92 <- climate_prcp_year_sum %>% 
  filter(year %in% 1959:1992)

climate_prcp_93_23 <- climate_prcp_year_sum %>% 
  filter(year %in% 1993:2023)
```

#### Rank-Sum Test

```{r}
prcp_91_58_rs <- wilcox.test(subset(climate_prcp_year_sum$total_all_prcp, climate_prcp_year_sum$year %in% 1891:1924),
                             subset(climate_prcp_year_sum$total_all_prcp, climate_prcp_year_sum$year %in% 1925:1958))

prcp_59_23_rs <- wilcox.test(subset(climate_prcp_year_sum$total_all_prcp, climate_prcp_year_sum$year %in% 1959:1992),
                             subset(climate_prcp_year_sum$total_all_prcp, climate_prcp_year_sum$year %in% 1993:2023))

prcp_91_23_rs <- wilcox.test(subset(climate_prcp_year_sum$total_all_prcp, climate_prcp_year_sum$year %in% 1891:1924),
                             subset(climate_prcp_year_sum$total_all_prcp, climate_prcp_year_sum$year %in% 1993:2023))
```

Comparing the years 1891-1924 *and* 1925-1958, the rank-sum test did yield statistically significant differences in the means with a p-value of 0.023.

Comparing the years 1959-1992 *and* 1993-2023, the rank-sum test did yield statistically significant differences in the means with a p-value of 1.72e-06.

Comparing the years 1891-1924 *and* 1993-2023, the rank-sum test did yield statistically significant differences in the means with a p-value of 4.13e-08.

## Temperature {.tabset}

```{r, include=FALSE}
# summary data
climate_temp_year_sum <- climate %>% 
  group_by(year) %>% 
  summarize(max_high_temp = max(tmax),
            min_high_temp = min(tmax),
            mean_high_temp = mean(tmax),
            max_low_temp = max(tmin),
            min_low_temp = min(tmin),
            mean_low_temp = mean(tmin)) %>% 
  mutate(year = as.numeric(year))
```

### Overall Time Period {.tabset}

#### Linear Model

```{r}
# graph plotting the average high and low temperatures each year with trend lines
# ggplot(climate_temp_year_sum) +
#   geom_point(aes(x = year, y = mean_high_temp), 
#              color = "red") +
#   geom_smooth(aes(x = year, y = mean_high_temp), 
#               formula = y ~ x, 
#               color = "black", 
#               method = "lm",
#               se = FALSE) +
#   geom_point(aes(x = year, y = mean_low_temp), 
#              color = "blue") +
#   geom_smooth(aes(x = year, y = mean_low_temp),
#               formula = y ~ x, 
#               color = "black",
#               method = "lm",
#               se = FALSE)

# trend line slopes
t_high_lm <- lm(mean_high_temp ~ year, data = climate_temp_year_sum)
t_high_confint <- confint(t_high_lm, "year", level = 0.95)

t_low_lm <- lm(mean_low_temp ~ year, data = climate_temp_year_sum)
t_low_confint <- confint(t_low_lm, "year", level = 0.95)

ggplot(data = climate_temp_year_sum) +
  stat_summary(aes(x = year, y = mean_high_temp),
               fun = "mean",
               geom = "point",
               color = "red") +
  stat_summary(aes(x = year, y = mean_low_temp),
               fun = "mean",
               geom = "point",
               color = "blue") +
  geom_smooth(aes(x = year, y = mean_high_temp),
              method = "lm",
              color = "darkred",
              se = FALSE) +
  geom_smooth(aes(x = year, y = mean_low_temp),
              method = "lm",
              color = "darkblue",
              se = FALSE) +
  labs(title = "Average High and Low Temperature from 1891-2023",
       x = "Year",
       y = "Temperature (°F)") +
  theme(plot.title = element_text(hjust = 0.5))
```

For average high temperatures, the y-intercept is `r round(t_high_lm$coefficients[1], 2)`. The slope for average high temperature per year is `r round(t_high_lm$coefficients[2], 2)` and is statistically significant with a p-value of 5.78e-05.

For average low temperatures, they y-intercept is `r round(t_low_lm$coefficients[1], 2)`. The slope for average low temperature per year is `r round(t_low_lm$coefficients[2], 2)` and is statistically significant with a p-value of 4.28e-05.

#### Mann-Kendall

```{r}
# Mann Kendall method for average high temperature
temp_high_mk <- MannKendall(climate_temp_year_sum$mean_high_temp)

# Mann Kendall method for average low temperature
temp_low_mk <- MannKendall(climate_temp_year_sum$mean_low_temp)
```

For average high temperatures, the strength of the trend is 0.32 and is statistically significant with a p-value of 2.22e-16.

For average low temperatures, the strength of the trend is 0.34 and is statistically significant with a p-value of 2.22e-16.

### Between Time Periods {.tabset}

```{r, include=FALSE}
climate_temp_91_24 <- climate_temp_year_sum %>% 
  filter(year %in% 1891:1924)

climate_temp_25_58 <- climate_temp_year_sum %>% 
  filter(year %in% 1925:1958)

climate_temp_59_92 <- climate_temp_year_sum %>% 
  filter(year %in% 1959:1992)

climate_temp_93_23 <- climate_temp_year_sum %>% 
  filter(year %in% 1993:2023)
```

#### Rank-Sum Test

```{r}
# Average High Temperature Rank-Sum Test
t_high_91_58_rs <- wilcox.test(subset(climate_temp_year_sum$mean_high_temp, climate_temp_year_sum$year %in% 1891:1924),
                               subset(climate_temp_year_sum$mean_high_temp, climate_temp_year_sum$year %in% 1925:1958))

t_high_59_23_rs <- wilcox.test(subset(climate_temp_year_sum$mean_high_temp, climate_temp_year_sum$year %in% 1959:1992),
                               subset(climate_temp_year_sum$mean_high_temp, climate_temp_year_sum$year %in% 1993:2023))

t_high_91_23_rs <- wilcox.test(subset(climate_temp_year_sum$mean_high_temp, climate_temp_year_sum$year %in% 1891:1924),
                               subset(climate_temp_year_sum$mean_high_temp, climate_temp_year_sum$year %in% 1993:2023))

# Average Low Temperature Rank-Sum Test
t_low_91_58_rs <- wilcox.test(subset(climate_temp_year_sum$mean_low_temp, climate_temp_year_sum$year %in% 1891:1924),
                              subset(climate_temp_year_sum$mean_low_temp, climate_temp_year_sum$year %in% 1925:1958))

t_low_59_23_rs <- wilcox.test(subset(climate_temp_year_sum$mean_low_temp, climate_temp_year_sum$year %in% 1959:1992),
                              subset(climate_temp_year_sum$mean_low_temp, climate_temp_year_sum$year %in% 1993:2023))

t_low_91_23_rs <- wilcox.test(subset(climate_temp_year_sum$mean_low_temp, climate_temp_year_sum$year %in% 1891:1924),
                              subset(climate_temp_year_sum$mean_low_temp, climate_temp_year_sum$year %in% 1993:2023))
```

For average high temperatures:

- Comparing the years 1891-1924 *and* 1925-1958, the rank-sum test did yield statistically significant differences in the means with a p-value of 0.0058.
- Comparing the years 1959-1992 *and* 1993-2023, the rank-sum test did not yield statistically significant differences in the means with a p-value of 0.18.
- Comparing the years 1891-1924 *and* 1993-2023, the rank-sum test did yield statistically significant differences in the means with a p-value of 1.36e-06.

For average low temperatures:

- Comparing the years 1891-1924 *and* 1925-1958, the rank-sum test did yield statistically significant differences in the means with a p-value of 2.44e-08.
- Comparing the years 1959-1992 *and* 1993-2023, the rank-sum test did not yield statistically significant differences in the means with a p-value of 0.69.
- Comparing the years 1891-1924 *and* 1993-2023, the rank-sum test did yield statistically significant differences in the means with a p-value of  1.26e-08.

# Seasonal Trends {.tabset}

## Precipitation (Rain and Snow) {.tabset}

```{r, include=FALSE}
climate_prcp_season <- climate %>% 
  group_by(year, season) %>% 
  summarize(most_rain = max(prcp),
            least_rain = min(prcp),
            total_rain = sum(prcp),
            most_snow = max(snow),
            least_snow = min(snow),
            total_snow = sum(snow),
            most_all_prcp = max(all_prcp),
            least_all_prcp = min(all_prcp),
            total_all_prcp = total_snow + total_rain) %>% 
  mutate(year = as.numeric(year))

clim_prcp_summer <- climate_prcp_season %>% 
  filter(season == "summer")

clim_prcp_winter <- climate_prcp_season %>% 
  filter(season == "winter")
```

### Summer {.tabset}

#### Overall Time Period {.tabset}

##### Linear Model

```{r}
# ggplot(clim_prcp_summer, 
#        aes(x = year, y = total_all_prcp)) +
#   geom_point(color = "blue") +
#   geom_smooth(formula = y ~ x,
#               color = "black",
#               method = "lm",
#               se = FALSE)

# trend line slope
prcp_summer_lm <- lm(total_all_prcp ~ year, data = clim_prcp_summer)
prcp_summer_confint <- confint(prcp_summer_lm, "year", level = 0.95)

ggplot(data = clim_prcp_summer,
       aes(x = year, y = total_all_prcp)) +
  stat_summary(fun = "mean", 
               geom = "point", 
               color = "blue") +
  geom_smooth(method = "lm",
              color = "black",
              se = FALSE)
```

For yearly total summer precipitation, the y-intercept is `r round(prcp_summer_lm$coefficients[1], 2)`. The slope of total summer precipitation by year is `r round(prcp_summer_lm$coefficients[2], 3)` and is not statistically significant (p-value = 0.069).

##### Mann-Kendall

```{r}
prcp_summer_mk <- MannKendall(clim_prcp_summer$total_all_prcp)
```

For yearly total summer precipitation, the strength is 0.206 and is statistically significant with a p-value of 0.00048.

#### Between Time Periods {.tabset}

```{r, include=FALSE}
clim_prcp_summer_91_24 <- clim_prcp_summer %>% 
  filter(year %in% 1891:1924)

clim_prcp_summer_25_58 <- clim_prcp_summer %>% 
  filter(year %in% 1925:1958)

clim_prcp_summer_59_92 <- clim_prcp_summer %>% 
  filter(year %in% 1959:1992)

clim_prcp_summer_93_23 <- clim_prcp_summer %>% 
  filter(year %in% 1993:2023)
```

##### Rank-Sum Test

```{r}
prcp_summer_91_58_rs <- wilcox.test(subset(clim_prcp_summer$total_all_prcp, clim_prcp_summer$year %in% 1891:1924),
                                    subset(clim_prcp_summer$total_all_prcp, clim_prcp_summer$year %in% 1925:1958))

prcp_summer_59_23_rs <- wilcox.test(subset(clim_prcp_summer$total_all_prcp, clim_prcp_summer$year %in% 1959:1992),
                                    subset(clim_prcp_summer$total_all_prcp, clim_prcp_summer$year %in% 1993:2023))

prcp_summer_91_23_rs <- wilcox.test(subset(clim_prcp_summer$total_all_prcp, clim_prcp_summer$year %in% 1891:1924),
                                    subset(clim_prcp_summer$total_all_prcp, clim_prcp_summer$year %in% 1993:2023))
```

Comparing the years 1891-1924 *and* 1925-1958, the rank-sum test did not yield statistically significant differences in the means with a p-value of 0.42.

Comparing the years 1959-1992 *and* 1993-2023, the rank-sum test did not yield statistically significant differences in the means with a p-value of 0.37.

Comparing the years 1891-1924 *and* 1993-2023, the rank-sum test did yield statistically significant differences in the means with a p-value of 0.01.

### Winter {.tabset}

#### Overall Time Period {.tabset}

##### Linear Model

```{r}
# ggplot(clim_prcp_winter, 
#        aes(x = year, y = total_all_prcp)) +
#   geom_point(color = "blue") +
#   geom_smooth(formula = y ~ x,
#               color = "black",
#               method = "lm",
#               se = FALSE)

# trend line slope
prcp_winter_lm <- lm(total_all_prcp ~ year, data = clim_prcp_winter)
prcp_winter_confint <- confint(prcp_winter_lm, "year", level = 0.95)

ggplot(data = clim_prcp_winter,
       aes(x = year, y = total_all_prcp)) +
  stat_summary(fun = "mean", 
               geom = "point", 
               color = "blue") +
  geom_smooth(method = "lm",
              color = "black",
              se = FALSE)
```

For yearly total winter precipitation, the y-intercept is `r round(prcp_winter_lm$coefficients[1], 2)`. The slope for yearly winter precipitation by year is `r round(prcp_winter_lm$coefficients[2], 2)` and is statistically significant with a p-value of 3.87e-06.

##### Mann-Kendall

```{r}
prcp_winter_mk <- MannKendall(clim_prcp_winter$total_all_prcp)
```

The strength of the trend is 0.311 and is statistically significant with a p-value of 1.19e-07.

#### Between Time Periods {.tabset}

```{r, include=FALSE}
clim_prcp_winter_91_24 <- clim_prcp_winter %>% 
  filter(year %in% 1891:1924)

clim_prcp_winter_25_58 <- clim_prcp_winter %>% 
  filter(year %in% 1925:1958)

clim_prcp_winter_59_92 <- clim_prcp_winter %>% 
  filter(year %in% 1959:1992)

clim_prcp_winter_93_23 <- clim_prcp_winter %>% 
  filter(year %in% 1993:2023)
```

##### Rank-Sum Test

```{r}
prcp_winter_91_58_rs <- wilcox.test(subset(clim_prcp_winter$total_all_prcp, clim_prcp_winter$year %in% 1891:1924),
                                    subset(clim_prcp_winter$total_all_prcp, clim_prcp_winter$year %in% 1925:1958))

prcp_winter_59_23_rs <- wilcox.test(subset(clim_prcp_winter$total_all_prcp, clim_prcp_winter$year %in% 1959:1992),
                                    subset(clim_prcp_winter$total_all_prcp, clim_prcp_winter$year %in% 1993:2023))

prcp_winter_91_23_rs <- wilcox.test(subset(clim_prcp_winter$total_all_prcp, clim_prcp_winter$year %in% 1891:1924),
                                    subset(clim_prcp_winter$total_all_prcp, clim_prcp_winter$year %in% 1993:2023))
```

Comparing the years 1891-1924 *and* 1925-1958, the rank-sum test did yield statistically significant differences in the means with a p-value of 0.019.

Comparing the years 1959-1992 *and* 1993-2023, the rank-sum test did yield statistically significant differences in the means with a p-value of 2.16e-07.

Comparing the years 1891-1924 *and* 1993-2023, the rank-sum test did yield statistically significant differences in the means with a p-value of 3.89e-08.

## Temperature {.tabset}

```{r, include=FALSE}
# summary data
climate_temp_season <- climate %>% 
  group_by(year, season) %>% 
  summarize(max_high_temp = max(tmax),
            min_high_temp = min(tmax),
            mean_high_temp = mean(tmax),
            max_low_temp = max(tmin),
            min_low_temp = min(tmin),
            mean_low_temp = mean(tmin)) %>% 
  mutate(year = as.numeric(year))

clim_temp_summer <- climate_temp_season %>% 
  filter(season == "summer")

clim_temp_winter <- climate_temp_season %>% 
  filter(season == "winter")
```

### Summer {.tabset}

#### Overall Time Period {.tabset}

##### Linear Model

```{r}
# graph plotting the average high and low temperatures each year with trend lines
# ggplot(clim_temp_summer) +
#   geom_point(aes(x = year, y = mean_high_temp), 
#              color = "red") +
#   geom_smooth(aes(x = year, y = mean_high_temp), 
#               formula = y ~ x, 
#               color = "black", 
#               method = "lm",
#               se = FALSE) +
#   geom_point(aes(x = year, y = mean_low_temp), 
#              color = "blue") +
#   geom_smooth(aes(x = year, y = mean_low_temp),
#               formula = y ~ x, 
#               color = "black",
#               method = "lm",
#               se = FALSE)

# trend line slopes
t_high_summer_lm <- lm(mean_high_temp ~ year, data = clim_temp_summer)
t_high_summer_confint <- confint(t_high_summer_lm, "year", level = 0.95)

t_low_summer_lm <- lm(mean_low_temp ~ year, data = clim_temp_summer)
t_low_summer_confint <- confint(t_low_summer_lm, "year", level = 0.95)

ggplot(data = clim_temp_summer, 
       aes(x = year, y = mean_low_temp)) +
  stat_summary(fun = "mean", 
               geom = "point", 
               color = "blue") +
  geom_smooth(method = "lm",
              color = "black",
              se = FALSE)

ggplot(data = clim_temp_summer) +
  stat_summary(aes(x = year, y = mean_high_temp),
               fun = "mean",
               geom = "point",
               color = "red") +
  stat_summary(aes(x = year, y = mean_low_temp),
               fun = "mean",
               geom = "point",
               color = "blue") +
  geom_smooth(aes(x = year, y = mean_high_temp),
              method = "lm",
              color = "darkred",
              se = FALSE) +
  geom_smooth(aes(x = year, y = mean_low_temp),
              method = "lm",
              color = "darkblue",
              se = FALSE)
```

For yearly summer average high temperatures, the y-intercept is `r round(t_high_summer_lm$coefficients[1], 2)`. The slope for average high temperature per year is `r round(t_high_summer_lm$coefficients[2], 3)` and is not statistically significant with a p-value of 0.066.

For yearly summer average low temperatures, they y-intercept is `r round(t_low_summer_lm$coefficients[1], 2)`. The slope for average low temperature per year is `r round(t_low_summer_lm$coefficients[2], 3)` and is statistically significant with a p-value of 6.01e-05.

##### Mann-Kendall

```{r}
# Mann Kendall method for average high temperature
t_high_summer_mk <- MannKendall(clim_temp_summer$mean_high_temp)

# Mann Kendall method for average low temperature
t_low_summer_mk <- MannKendall(clim_temp_summer$mean_low_temp)
```

For yearly summer average high temperatures, the strength of the trend is 0.12 and is statistically significant with a p-value of 0.04.

For yearly summer average low temperatures, the strength of the trend is 0.25 and is statistically significant with a p-value of 3.52e-05.

#### Between Time Periods {.tabset}

```{r, include=FALSE}
climate_temp_summer_91_24 <- clim_temp_summer %>% 
  filter(year %in% 1891:1924)

climate_temp_summer_25_58 <- clim_temp_summer %>% 
  filter(year %in% 1925:1958)

climate_temp_summer_59_92 <- clim_temp_summer %>% 
  filter(year %in% 1959:1992)

climate_temp_summer_93_23 <- clim_temp_summer %>% 
  filter(year %in% 1993:2023)
```

##### Rank-Sum Test

```{r}
# Average High Temperature Rank-Sum Test
t_high_summer_91_58_rs <- wilcox.test(subset(clim_temp_summer$mean_high_temp, clim_temp_summer$year %in% 1891:1924),
                                      subset(clim_temp_summer$mean_high_temp, clim_temp_summer$year %in% 1925:1958))

t_high_summer_59_23_rs <- wilcox.test(subset(clim_temp_summer$mean_high_temp, clim_temp_summer$year %in% 1959:1992),
                                      subset(clim_temp_summer$mean_high_temp, clim_temp_summer$year %in% 1993:2023))

t_high_summer_91_23_rs <- wilcox.test(subset(clim_temp_summer$mean_high_temp, clim_temp_summer$year %in% 1891:1924),
                                      subset(clim_temp_summer$mean_high_temp, clim_temp_summer$year %in% 1993:2023))

# Average Low Temperature Rank-Sum Test
t_low_summer_91_58_rs <- wilcox.test(subset(clim_temp_summer$mean_low_temp, clim_temp_summer$year %in% 1891:1924),
                                     subset(clim_temp_summer$mean_low_temp, clim_temp_summer$year %in% 1925:1958))

t_low_summer_59_23_rs <- wilcox.test(subset(clim_temp_summer$mean_low_temp, clim_temp_summer$year %in% 1959:1992),
                                     subset(clim_temp_summer$mean_low_temp, clim_temp_summer$year %in% 1993:2023))

t_low_summer_91_23_rs <- wilcox.test(subset(clim_temp_summer$mean_low_temp, clim_temp_summer$year %in% 1891:1924),
                                     subset(clim_temp_summer$mean_low_temp, clim_temp_summer$year %in% 1993:2023))
```

For yearly average summer high temperatures:

- Comparing the years 1891-1924 *and* 1925-1958, the rank-sum test did yield not statistically significant differences in the means with a p-value of 0.36.
- Comparing the years 1959-1992 *and* 1993-2023, the rank-sum test did not yield statistically significant differences in the means with a p-value of 0.16.
- Comparing the years 1891-1924 *and* 1993-2023, the rank-sum test did yield statistically significant differences in the means with a p-value of 0.018.

For yearly average summer low temperatures:

- Comparing the years 1891-1924 *and* 1925-1958, the rank-sum test did yield statistically significant differences in the means with a p-value of 1.6e-05.
- Comparing the years 1959-1992 *and* 1993-2023, the rank-sum test did not yield statistically significant differences in the means with a p-value of 0.7.
- Comparing the years 1891-1924 *and* 1993-2023, the rank-sum test did yield statistically significant differences in the means with a p-value of 4.74e-06.

### Winter {.tabset}

#### Overall Time Period {.tabset}

##### Linear Model

```{r}
# graph plotting the average high and low temperatures each year with trend lines
# ggplot(clim_temp_winter) +
#   geom_point(aes(x = year, y = mean_high_temp), 
#              color = "red") +
#   geom_smooth(aes(x = year, y = mean_high_temp), 
#               formula = y ~ x, 
#               color = "black", 
#               method = "lm",
#               se = FALSE) +
#   geom_point(aes(x = year, y = mean_low_temp), 
#              color = "blue") +
#   geom_smooth(aes(x = year, y = mean_low_temp),
#               formula = y ~ x, 
#               color = "black",
#               method = "lm",
#               se = FALSE)

# trend line slopes
t_high_winter_lm <- lm(mean_high_temp ~ year, data = clim_temp_winter)
t_high_winter_confint <- confint(t_high_winter_lm, "year", level = 0.95)

t_low_winter_lm <- lm(mean_low_temp ~ year, data = clim_temp_winter)
t_low_winter_confint <- confint(t_low_winter_lm, "year", level = 0.95)

ggplot(data = clim_temp_winter) +
  stat_summary(aes(x = year, y = mean_high_temp),
               fun = "mean",
               geom = "point",
               color = "red") +
  stat_summary(aes(x = year, y = mean_low_temp),
               fun = "mean",
               geom = "point",
               color = "blue") +
  geom_smooth(aes(x = year, y = mean_high_temp),
              method = "lm",
              color = "darkred",
              se = FALSE) +
  geom_smooth(aes(x = year, y = mean_low_temp),
              method = "lm",
              color = "darkblue",
              se = FALSE)
```

For yearly winter average high temperatures, the y-intercept is `r round(t_high_winter_lm$coefficients[1], 2)`. The slope for average high temperature per year is `r round(t_high_winter_lm$coefficients[2], 3)` and is statistically significant with a p-value of 2.33e-06.

For yearly winter average low temperatures, they y-intercept is `r round(t_low_winter_lm$coefficients[1], 2)`. The slope for average low temperature per year is `r round(t_low_winter_lm$coefficients[2], 3)` and is statistically significant with a p-value of 5.6e-05.

##### Mann-Kendall

```{r}
# Mann Kendall method for average high temperature
t_high_winter_mk <- MannKendall(clim_temp_winter$mean_high_temp)

# Mann Kendall method for average low temperature
t_low_winter_mk <- MannKendall(clim_temp_winter$mean_low_temp)
```

For yearly winter average high temperatures, the strength of the trend is 0.26 and is statistically significant with a p-value of 1.19e-05.

For yearly winter average low temperatures, the strength of the trend is 0.21 and is statistically significant with a p-value of 0.00029.

#### Between Time Periods {.tabset}

```{r, include=FALSE}
climate_temp_winter_91_24 <- clim_temp_winter %>% 
  filter(year %in% 1891:1924)

climate_temp_winter_25_58 <- clim_temp_winter %>% 
  filter(year %in% 1925:1958)

climate_temp_winter_59_92 <- clim_temp_winter %>% 
  filter(year %in% 1959:1992)

climate_temp_winter_93_23 <- clim_temp_winter %>% 
  filter(year %in% 1993:2023)
```

##### Rank-Sum Test

```{r}
# Average High Temperature Rank-Sum Test
t_high_winter_91_58_rs <- wilcox.test(subset(clim_temp_winter$mean_high_temp, clim_temp_winter$year %in% 1891:1924),
                                      subset(clim_temp_winter$mean_high_temp, clim_temp_winter$year %in% 1925:1958))

t_high_winter_59_23_rs <- wilcox.test(subset(clim_temp_winter$mean_high_temp, clim_temp_winter$year %in% 1959:1992),
                                      subset(clim_temp_winter$mean_high_temp, clim_temp_winter$year %in% 1993:2023))

t_high_winter_91_23_rs <- wilcox.test(subset(clim_temp_winter$mean_high_temp, clim_temp_winter$year %in% 1891:1924),
                                      subset(clim_temp_winter$mean_high_temp, clim_temp_winter$year %in% 1993:2023))

# Average Low Temperature Rank-Sum Test
t_low_winter_91_58_rs <- wilcox.test(subset(clim_temp_winter$mean_low_temp, clim_temp_winter$year %in% 1891:1924),
                                     subset(clim_temp_winter$mean_low_temp, clim_temp_winter$year %in% 1925:1958))

t_low_winter_59_23_rs <- wilcox.test(subset(clim_temp_winter$mean_low_temp, clim_temp_winter$year %in% 1959:1992),
                                     subset(clim_temp_winter$mean_low_temp, clim_temp_winter$year %in% 1993:2023))

t_low_winter_91_23_rs <- wilcox.test(subset(clim_temp_winter$mean_low_temp, clim_temp_winter$year %in% 1891:1924),
                                     subset(clim_temp_winter$mean_low_temp, clim_temp_winter$year %in% 1993:2023))
```

For yearly average winter high temperatures:

- Comparing the years 1891-1924 *and* 1925-1958, the rank-sum test did yield statistically significant differences in the means with a p-value of 0.00063.
- Comparing the years 1959-1992 *and* 1993-2023, the rank-sum test did yield statistically significant differences in the means with a p-value of 0.032.
- Comparing the years 1891-1924 *and* 1993-2023, the rank-sum test did yield statistically significant differences in the means with a p-value of 7.83e-05.

For yearly average winter low temperatures:

- Comparing the years 1891-1924 *and* 1925-1958, the rank-sum test did yield statistically significant differences in the means with a p-value of 7.91e-05.
- Comparing the years 1959-1992 *and* 1993-2023, the rank-sum test did not yield statistically significant differences in the means with a p-value of 0.14.
- Comparing the years 1891-1924 *and* 1993-2023, the rank-sum test did yield statistically significant differences in the means with a p-value of 0.00021.

# Data Citation

NOAA. (2023). Daily Summaries at the Ann Arbor University of Michigan Station. NOAA. https://www.ncei.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00200230/detail